---
title: "Exercises Day 1 Part 2"
author: "Thomas Kuo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the dataset

You use the `##` to create a header. Use the header to indicate the question you are answering. After the header, in this space here, annotate your answer, telling us what you did.

```{r}
# Here is where your R code goes. ``` opens the code chunk, followed by {r}. Then ``` closes the chunk. When you 'knit' the file, the codes will evaluate.

gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```

## If Statement
Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the any function.
```{r Testing Years For If}
test_year <- 2002
if (any(gapminder$year == test_year)) {
    print(paste0("There is data for ", test_year))
} else {
        print(paste0("There are no records in ",test_year))
}

test_year <- 2012
if (any(gapminder$year == test_year)) {
    print(paste0("There is data for ", test_year))
} else {
        print(paste0("There are no records in ",test_year))
}
```
## Loop and If Statements
Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.
```{r Loop and If}
dataset_mean <- mean(gapminder$pop)
for (i in unique(gapminder$country)) {
    temp_cpop <- mean(gapminder$pop[gapminder$country == i]) #country population
    temp_cmLE <- mean(gapminder$lifeExp[gapminder$country == i]) #mean country life expectancy
    
    if (temp_cpop < dataset_mean) {
        print(paste0("The mean life expectancy for ", i, " is ", temp_cmLE))
    } else {
        print(paste0("The population for ", i, " is higher than the mean population of the data set"))
    }
}
```


Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.
```{r Loop and If - Part 2}
for (i in unique(gapminder$continent)){
    temp_cmLE <- mean(gapminder$lifeExp[gapminder$continent == i]) #mean continent life expectancy
    
    if (temp_cmLE < 50) {
        print(paste0("The mean life expectancy for ", i, " is ", round(temp_cmLE), " and is less than 50"))
    } else {
        if (temp_cmLE > 70) {
        print(paste0("The mean life expectancy for ", i, " is ", round(temp_cmLE), " and is greater than 70"))
    } else {
        print(paste0("The mean life expectancy for ", i, " is ", round(temp_cmLE), " and is between 50 & 70"))
    }
    }
}
```

## Exercise: Writing Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.

```{r Writing Functions}
print_col_and_class <-
    function(dataframe) {
        my_names <- names(dataframe)
        my_class <- lapply(dataframe, class)
        
        ##length gets the numeric length of my_names
        ##seq turns that length into a sequence to iterate that many times
        for (i in seq(length(my_names))) {
        print(paste0("Column: ", my_names[i], " || Data Class: ", my_class[i]))
        }
    }

print_col_and_class(gapminder)
```


Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.

```{r Writing Functions - Part 2}
print_sd_med <-
    function(vector, boolean = FALSE) {
        fx_mean <- mean(vector)
        fx_sd <- sd(vector)
        
        #use cat when you have newline character, new line does not work with print
        cat(paste0("Mean: ", fx_mean, "\nStandard Deviation: ", fx_sd))
        if (boolean == TRUE) {
            cat(paste0("\nMedian: ", median(vector),"\n"))
        } 
 }

print_sd_med(gapminder$lifeExp, TRUE)
print_sd_med(gapminder$lifeExp,)
```

## Analyzing the Relationship

Use what you???ve learned so far to answer the following questions using the gapminder dataset. Be sure to include some visualizations!

1/ What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r Analyzing Relationship}
gapminder$log_gdp <- log(gapminder$gdpPercap)
gapminder$log_le <- log(gapminder$lifeExp)
    
years <- unique(gapminder$year)
corr_vec <- NULL
    for (i in years) {
        year_corr <- cor(gapminder$log_gdp[gapminder$year == i],gapminder$log_le[gapminder$year == i])
        corr_vec <- c(corr_vec, year_corr)
    }

plot(corr_vec ~ years, main = "Correlation of GDP & Life Expectancy by Year", xlab = "Year", ylab = "Correlation")
```


2/ Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r Analyzing Relationships - Part 2}
##Split Americas into North and South
gapminder$cont_affinitized <- gapminder$continent
gapminder$cont_affinitized[gapminder$cont_affinitized == "Americas"] <- NA
gapminder$cont_affinitized[gapminder$country == "Canada"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Guatemala"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Costa Rica"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Cuba"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Dominican Republic"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "El Salvador"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Haiti"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Honduras"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Jamaica"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Nicaragua"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Panama"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "Trinidad and Tobago"] <- "North America"
gapminder$cont_affinitized[gapminder$country == "United States"] <- "North America"
gapminder$cont_affinitized[is.na(gapminder$cont_affinitized)] <- "South America"

##evaluate relationship between GDP and life expectancy
gapminder$log_gdp <- log(gapminder$gdpPercap)
gapminder$log_le <- log(gapminder$lifeExp)
    
continents <- unique(gapminder$cont_affinitized)
corr_vec <- NULL
    for (i in continents) {
        cont_corr <- cor(gapminder$log_gdp[gapminder$cont_affinitized == i],gapminder$log_le[gapminder$cont_affinitized == i])
        corr_vec <- c(corr_vec, cont_corr)
    }
my_df <- data.frame(continents, corr_vec)
library(ggplot2)
ggplot(data = my_df, aes(x = continents, y = corr_vec)) + geom_bar(stat="identity") + ggtitle("Correlation of GDP per capita and Life Expectancy by Continent") + labs(x = "Continent", y = "Correlation")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
