---
title: "R Day 2 exercises"
author: "Thomas Kuo"
date: "`r Sys.Date()`"
output: html_document
---

```{r global_options, echo = FALSE, include = FALSE}

```

<!-- BEFORE YOU START, REMEMBER TO RENAME THE .RMD FILE AND SAVE IT AS day2_exercises_LastNameFirstName.Rmd AND THEN REMOVE THIS INSTRUCTION -->

## Load the data

Read both California energy datasets. Make sure the `datetime` variable is in an appropriate data type (i.e. not character). 

```{r}
library(here)
library(readr)
library(lubridate)
library(dplyr)
library(reshape2)
generation <- read.csv(here::here("data/ca_energy_generation.csv"), stringsAsFactors = FALSE)
imports <- read.csv(here::here("data/ca_energy_imports.csv"), stringsAsFactors = FALSE)
generation$datetime <- as_datetime(generation$datetime)
imports$datetime <- as_datetime(imports$datetime)
```


## Merge and reshape the data

Merge the two datasets and then melt the resulting dataframe/datatable to make it tidy.

```{r}
merged_energy <- merge(generation, imports, by = "datetime")
long_me <- melt(merged_energy, id.vars = "datetime",
                variable.name = "source", #the "column" name becomes the variable name
                value.name = "usage") #the values are the values stored within the column
```


## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Bonus: If you are using `dplyr`, try to do this all in one pipe!

```{r}
long_me <- long_me %>% ##adding columns to the long_me data frame
    mutate(day = as_date(datetime), #adding the variables "day" and "log_output"
           log_output = log(usage)) %>%
    group_by(day) %>% #grouping by day for total daily output
    mutate(total_daily_output = sum(usage, na.rm = TRUE)) %>% 
    ungroup() %>% #ungrouping once daily outputs are assigned to each entry in a given day
    mutate(per_output = usage / total_daily_output) #divide usage by total daily output for percent
```


## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)

```{r Q3P1}
#1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
library(dplyr)
long_me %>%
    group_by(source) %>% #group by source to consolidate all usage by source
    summarize(hourly_mean = mean(usage), na.rm = T) %>% #summarize the mean output into a new variable
    arrange(desc(hourly_mean)) #arrange in terms of descending

## natural_gas has the most, while coal has the least
```

```{r Q3P2}
#2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
long_me %>%
    filter(usage > 0) %>%
    group_by(source, day) %>%
    summarize(daily_mean = mean(usage), na.rm = T) %>%
    arrange(desc(daily_mean)) %>%
    head(5)
#natural gas has the greatest mean output by day on 2019-09-05

long_me %>%
    filter(usage > 0) %>%
    group_by(source, day) %>%
    summarize(daily_mean = mean(usage), na.rm = T) %>%
    arrange(desc(daily_mean)) %>%
    tail(5)
#coal has the least mean output by day on 2019-09-07
```

```{r Q3P3}
#3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)
long_me %>%
    filter(usage > 0) %>%
    group_by(source) %>%
    summarize(source_var = sd(usage), na.rm = T) %>%
    arrange(desc(source_var))

#natural_gas has the greatest variance
#coal has the smallest variance
```


## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with yourdata manipulation skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by?

```{r}
regroup_data <- read.csv(here::here("data/ca_energy_regroup.csv"), stringsAsFactors = FALSE)
long_me <- left_join(long_me, regroup_data, by = c("source" = "type"))
```

